Unsupervised embedding of trajectories captures the latent structure of mobility

### 2021-01-30 paper/rebuttal/rebuttal.tex

\response{
\todo{I'm worry this sounds too dismissive. Edit please?}
Thank you for this comment. 
We feel that applying these embeddings to other areas of scientific activity would be incredibly enlightening! 
Your recommendation of collaboration (based on the author byline) and advisor-student pairs are especially interesting, and we would aim to address these areas in future studies. 
However, given the importance of mobility to science, the importance of human mobility more generally, and the connection we discovered between \textit{word2vec} and the gravity model, that a focus of co-affiliation, which most closely aligns with ideas of mobility, is the most reasonable data to use for this manuscript. 
}

\rcomment{
There are known concerns with clustering a 300 dimensional objects because of the curse of
dimensionality. Many have shown that reducing dimensionality with UMAP to a nontrivial
number of dimensions (e.g., 30), which preserves both local and some global distance, then
clustering on those dimensions are much more stable (e.g., this post). This was one
concern with the stability of the clustering analysis

}

\response{
\todo{will test. But what do we test exactly?}
\todo{JS: actually, can't agree with this opinion. For sure, curse of  conventionality exists. Many dimensional reduction technique do is reducing dimension with keeping pair-wise distance. So, it can be well matched with some clustering algorithm such as k-means algorithm. But, in our analysis, we used raw similarity score for hierarchical clustering. So, in this analysis, using projected similarity form umap(their suggestion) is kinda ridiculous. Approximation is approximation, not real similarity.
}
}

### 2021-01-30 paper/rebuttal/rebuttal.tex

The authors demonstrate three key findings, with their empirical analysis based on datasets of flight itineraries, academic career trajectories, and Korean accommodation reservations.
The authors show that the distance implicitly defined by the embedding space better reflects the flux between locations, compared to geographic distance (and personalized PageRank, in the appendix). More concretely, the authors show that the embedding distance explains more variance for the flow when using a gravity model approach.
The authors show how the “global” structure of the embedding space can be visualized and interpreted by using UMAP to generate two-dimensional visualizations. Notably, the authors demonstrate interesting and non-trivial organization of the data at multiple scales, based on the embeddings.
Finally, the authors show that the embeddings encode a latent “prestige” hierarchy over the academic data (e.g., with the L2 norm having an interesting non-linear relationship with institution size that appears to be mediated by prestige).
}

\response{%
Thank you for your valuable comments. 
In the following, we address each of your comments point-by-point.
}

\rcomment{%
To the best of my knowledge, this is the first work providing a thorough analysis of the structure encoded by latent mobility embeddings (e.g., the prestige hierarchy), and this is the first work I am aware of to propose embeddings as an explanatory/exploratory tool in this context. In this way, the work has some originality and significance.

\begin{itemize}
	\item Liu, Xin, Yong Liu, and Xiaoli Li. "Exploring the Context of Locations for Personalized Location Recommendations." IJCAI. 2016.
\item Feng, Shanshan, et al. "POI2Vec: geographical latent representation for predicting future visitors." AAAI. 2017.
\item Yao, Zijun, et al. "Representing Urban Functions through Zone Embedding with Human Mobility Patterns." IJCAI. 2018.
\item Cao, Hancheng, et al. "Habit2vec: Trajectory semantic embedding for living pattern recognition in population." IEEE Transactions on Mobile Computing 19.5 (2019): 1096-1108.
\end{itemize}

@@

\response{%
Thank you for pointing this out. 
TODO:
We carefully read out the Barthélemy, Physics Reports 499, 1, 2011 and other reference paper in draft.
Also, we also report the performance of the flux estimation with different deterrence functions, which shows Embedding distance with exponential decay function always produce the best result.

Beyond the ﬁtting on the data, we would like to argue that geographical distance ﬁts well with the power-law decay function because it is is more suitable for analyzing large, complex, and scale-free spatial systems , while embedding distance ﬁts well with the exponential decay function, which we will show later in the section of mathematical relation between word2vecand the gravity model.
}

\rcomment{
I would suggest to the authors to carefully cite other papers applying Word2Vec embedding on spatial data, especially focusing on location embedding based on trajectory sequences. Examples are:
\begin{itemize}
\item A. Solomon et al., Predict Demographic Information Using Word2Vec on Spatial Trajectories., ACM UMAP '18: Proceedings of the 26th Conference on User Modeling, Adaptation and Personalization, July 2018 Pages 331–339, https://doi.org/10.1145/3209219.3209224
\item A. Crivellari and E. Beinat, From Motion Activity to Geo-Embeddings: Generating and Exploring Vector Representations of Locations, Traces and Visitors through Large-Scale Mobility Data., ISPRS Int. J. Geo-Inf. 2019, 8(3), 134 (2019); https://doi.org/10.3390/ijgi8030134

\end{itemize}

The application of the Word2Vec method on the studied trajectory sequences is not well explained. Without explaining the embedding method briefly in the main text the reader has no idea how to interpret the embedding distance, which is in the hear of the study.

### 2021-01-30 paper/current/main.tex
update main text with YY comments

\begin{equation}
	\label{eq:gravity_basic}
	\hat{T_{ij}} = Cm_{i}m_{j}f(r_{ij}),
\end{equation}
where $m_{i}$ is the population of locations $i$,  $f(r_{ij})$ is a decay function with respect to distance between locations, and $C$ is a constant estimated from data (see Methods).
For the flight itinerary data, we use population $m_{i}$ as the total number of unique passengers who passed through each airport, for the Korean accommodation reservation data, we use the total number of unique customers who booked with each accommodation, and for scientific mobility, we use the mean annual number of unique mobile and non-mobile authors who were affiliated with each organization.
$\hat{T}_{ij}$, which is often referred to as ``expected flux''~\autocite{simini2012universal}, is the expected frequency of the co-occurrence of location $i$ and $j$ in the trajectory in the gravity model.

@@

The {\it word2vec} model takes a location trajectory, denoted by ($a_{1}, a_{2}, \ldots, a_{T}$), as an input.
A location $a_t=i$ is considered to have a context location $a_{t'}=j$ that appears in its succeeding or proceeding $w$ locations in the trajectory, \ie  $j \in [a_{t-w}, \ldots, a_{t-1}, a_{t+1},\ldots, a_{t+w}]$.
 {\it Word2vec} learns an embedding by estimating the probability that location $i$ has context $j$:
\begin{align}
	P\left(j \given i \right):= \frac{\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i}, \label{eq:prob_w2v_ng}
\end{align}
where the denominator $Z_i=\sum_{j' \in \mathcal{A}} \exp(\bm{u}_{j'} \cdot \bm{v}_{i})$ is a normalization constant, and $\mathcal{A}$ is the set of all locations.
Although {\it word2vec} generates two embedding vectors $\bm{v_i}$ and $\bm{u}_i$---which are referred to as the in-vector and out-vector, respectively, we follow many studies by adopting the in-vector $\bm{v}_i$ as an embedding of location $i$.

Our experiments show that {\it word2vec} best explains real-world mobility flow when $w=1$ (Fig.~\ref{fig:supp:hyperparameter}), with the flow predicted by
\begin{align}
	\label{eq:flow_w2v}
	\hat{T_{ij}}= P(i) P(j\given i) = \frac{P(i)\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i},
\end{align}
where $P(i)$ is the probability that location $i$ appears in the trajectory.
Because computing $Z_i$ is computationally demanding, multiple approximation methods have been proposed, the most  popular being negative sampling~\autocite{mikolov2013word2vec}. 
Negative sampling finds the embedding vectors by solving a binary classification task, in which the model classifies whether a given center-context location pair $(i, j)$ comes from the data or artificially generated ``negative'' samples.
The negative pair $(i,j)$ is sampled in proportion to $P^\gamma(j)$ with parameter $\gamma \geq 0$.

@@

Parameter $\gamma=1$ is a special choice that ensures that, when the embedding dimension is sufficiently large, there exists optimial in-vectors and out-vectors such that $\bm{u}_i = \bm{v}_i$ \autocite{levy2014neural}.
Setting $\gamma = 1$ and substituting $\bm{u}_i = \bm{v}_j$ into Eq.~\eqref{eq:prob_w2v_ng0}, the flow predicted by {\it word2vec} is given by
\begin{align}
	\hat{T_{ij}}= P(i) P(j\given i) = \frac{P(i)P(j)\exp(\bm{v}_j \cdot \bm{v}_{i})}{Z_i}. \label{eq:flow_w2v_ng1}
\end{align}
The flow $\hat{T_{ij}}$ is symmetric (\ie $\hat{T_{ij}}=\hat{T_{ji}}$) because {\it word2vec} neglects whether the context $j$ appears before or after the center $i$ in the trajectory.
If we swap $i$ and $j$ in Eq.~\eqref{eq:flow_w2v_ng1}, the numerator remains the same but the denominator can be different.
Therefore, to ensure $T_{ij} = T_{ji}$, the denominator $Z_i$ should be a constant.

Taken altogether, the {\it word2vec} model with the negative sampling predicts a flow in the same form as in the gravity model:
\begin{align}
	\hat{T_{ij}}=  C P(i) P(j)  \exp(\bm{v}_j \cdot \bm{v}_{i}), \label{eq:flow_w2v_ng0}
\end{align}
where $C$ is a positive constant.
In other words, \emph{word2vec with the negative sampling is a gravity model}, with the mass given by the location's frequency $P(i)$, and the distance measured by their dot similarities.
While the gravity model generates the flow from the given mass and locations, {\it word2vec} finds the mass and locations that best explain the given trajectories.
Therefore, the embedding distance is inherently tied to the flow, and hence, has greater predictive power than the geographic distance.

@@

Negative sampling is a biased estimator~\autocite{Chia2010,Dyer2014} that optimizes a distinct model  from {\it word2vec}.
What is the different model?

Negative sampling trains {\it word2vec} using a binary classification task as follows.
A center-context word pair $(i,j)$ is sampled from the given data as a positive pair, and $k$ negative pairs $(i,\ell)$ are sampled from a noise distribution $p_0(\ell)=P^\gamma(\ell)$. ($P(j)$ is the fraction of $j$ in the data, and $\gamma>0$ is a parameter).
Then, given a center-context word pair $(i,j)$, the model predicts whether it is a positive or negative sample using sigmoid function:
\begin{align}
	\label{eq:sigmoid}
	\frac{1}{1 + \exp\left( - \bm{u}_j \cdot \bm{v}_i \right)}.
\end{align}

This procedure does not guarantee that the embedding optimally converges, even when increasing the training samples and iterations~\autocite{Chia 2010,Dyer2014}.
This problem is addressed by the noise contrastive estimation (NCE) \autocite{Chia2010}, which fits a probability model in the form:
\begin{align}
	\label{eq:nce}
	f(x) / \sum_{x'} f(x'),
\end{align}
where $f(x)$ is the unnormalized probability for an instance $x$.
As in negative sampling, NCE fits the model based on a binary classification task, with positive and negative samples randomly drawn from the given data and a noise distribution $p_0$, respectively, using a bias-corrected sigmoid function:
\begin{align}
	\label{eq:sigmoid2}
	\frac{1}{1 + \exp\left( -f(x) + \log p_0(x) \right)}.
\end{align}
In case of the {\it word2vec}, the instance is the dot similarity $x=\bm{u}_i \cdot \bm{v}_j$,  $f(x)=\exp(x)$, and $p_0(j)=P^\gamma(j)$.

@@

We model co-occurences $T_{ij}$ for locations $i$ and $j$ (referred to as flux), using the gravity law of mobility~\autocite{zipf1946gravity}.
The gravity law of mobility, which was inspired by Newton's law of gravity, postulates that attraction between two locations is a function of their population and the distance between them.
This formulation and variants have proven useful for modeling and predicting many kinds of mobility~\autocite{jung2008highwaygravity, curiel2018citygravity, truscott2012epidemicgravity, hong2016busgravity}.
In the gravity law of mobility, the \textit{expected flux}, $\hat{T_{ij}}$ between two locations $i$ and $j$ is defined as,

%
% Gravity law EQ
%
\begin{equation}
	\label{eq:gravity_basic}
	\hat{T_{ij}} = Cm_{i}m_{j}f(r_{ij}),
\end{equation}
where $m_{i}$ and $m_{j}$ are the population of locations, defined as the total number of passenger who passed through each airport for U.S. airport itineraries,  the total number of customer who booked with each accommodation for Korean accommodation reservations, and the yearly-average count of unique authors, both mobile and non-mobile, affiliated with each organization for scientific mobility.
$f(r_{ij})$ is a decay function of distance $r_{ij}$ between locations $i$ and $j$.
Here, we used the most basic gravity model which assumes symmetry of the flow $\hat{T_{ij}} =\hat{T_{ji}}$ and distance $r_{ij} = r_{ji}$, while there are four proposed variants \autocite{wilson2011entropy}.
There are two popular forms for the $f(r_{ij})$:
one is a power law function in the form $f(r_{ij})= r^{-\alpha}_{ij} \ (\alpha > 0)$, and the other is an exponential function in the form $f(r_{ij}) = e^{-\beta r_{ij}} \ (\beta > 0)$ \autocite{chen2015distance}.
The parameters for $f(r_{ij})$ and $C$ are fit to given mobility data using a log-linear regression~\autocite{jung2008highwaygravity, curiel2018citygravity, truscott2012epidemicgravity, hong2016busgravity, simini2012universal}.

>>>>

\begin{equation}
	\label{eq:gravity_basic}
	\hat{T}_{ij} = Cm_{i}m_{j}f(r_{ij}),
\end{equation}
where $m_{i}$ is the population of locations $i$,  $f(r_{ij})$ is a decay function with respect to distance between locations, and $C$ is a constant estimated from data (see Methods).
For the flight itinerary data, we use population $m_{i}$ as the total number of unique passengers who passed through each airport, for the Korean accommodation reservation data, we use the total number of unique customers who booked with each accommodation, and for scientific mobility, we use the mean annual number of unique mobile and non-mobile authors who were affiliated with each organization.
$\hat{T}_{ij}$, which is often referred to as ``expected flux''~\autocite{simini2012universal}, is the expected frequency of the co-occurrence of location $i$ and $j$ in the trajectory in the gravity model.

@@

The {\it word2vec} model takes a location trajectory, denoted by ($a_{1}, a_{2}, \ldots, a_{T}$), as an input.
A location $a_t=i$ is considered to have a context location $a_{t'}=j$ that appears in its succeeding or proceeding $w$ locations in the trajectory, \ie  $j \in [a_{t-w}, \ldots, a_{t-1}, a_{t+1},\ldots, a_{t+w}]$.
 {\it Word2vec} learns an embedding by estimating the probability that location $i$ has context $j$:
\begin{align}
	P\left(j \given i \right):= \frac{\exp(\bm{u}_j \cdot \bm{v}_{i})}{\widetilde{Z_i}}, \label{eq:prob_w2v_ng}
\end{align}
where the denominator $\widetilde{Z_i}=\sum_{j' \in \mathcal{A}} \exp(\bm{u}_{j'} \cdot \bm{v}_{i})$ is a normalization constant, and $\mathcal{A}$ is the set of all locations.
Although {\it word2vec} generates two embedding vectors $\bm{v_i}$ and $\bm{u}_i$---which are referred to as the in-vector and out-vector, respectively, we follow many studies by adopting the in-vector $\bm{v}_i$ as an embedding of location $i$.

Our experiments show that {\it word2vec} best explains real-world mobility flow when $w=1$ (Fig.~\ref{fig:supp:hyperparameter}), with the flow predicted by
\begin{align}
	\label{eq:flow_w2v}
	\hat{T}_{ij} = P(i) P(j\given i) = \frac{P(i)\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i},
\end{align}
where $P(i)$ is the probability that location $i$ appears in the trajectory.
Because computing $Z_i$ is computationally demanding, multiple approximation methods have been proposed, the most  popular being negative sampling~\autocite{mikolov2013word2vec}. 
Negative sampling finds the embedding vectors by solving a binary classification task, in which the model classifies whether a given center-context location pair $(i, j)$ comes from the data or artificially generated ``negative'' samples.
The negative pair $(i,j)$ is sampled in proportion to $P^\gamma(j)$ with parameter $\gamma \geq 0$.

@@

Parameter $\gamma=1$ is a special choice that ensures that, when the embedding dimension is sufficiently large, there exists optimial in-vectors and out-vectors such that $\bm{u}_i = \bm{v}_i$ \autocite{levy2014neural}.
Setting $\gamma = 1$ and substituting $\bm{u}_i = \bm{v}_j$ into Eq.~\eqref{eq:prob_w2v_ng0}, the flow predicted by {\it word2vec} is given by
\begin{align}
	\hat{T_{ij}}= P(i) P(j\given i) = \frac{P(i)P(j)\exp(\bm{v}_j \cdot \bm{v}_{i})}{Z_i}. \label{eq:flow_w2v_ng1}
\end{align}
The flow $\hat{T}_{ij}$ is symmetric (\ie $\hat{T}_{ij}=\hat{T}_{ji}$) because {\it word2vec} neglects whether the context $j$ appears before or after the center $i$ in the trajectory.
If we swap $i$ and $j$ in Eq.~\eqref{eq:flow_w2v_ng1}, the numerator remains the same but the denominator can be different.
Therefore, to ensure $T_{ij} = T_{ji}$, the denominator $Z_i$ must be a constant.

Taken altogether, the {\it word2vec} model with the negative sampling predicts a flow in the same form as in the gravity model:
\begin{align}
	\hat{T}_{ij}=  C P(i) P(j)  \exp(\bm{v}_j \cdot \bm{v}_{i}), \label{eq:flow_w2v_ng0}
\end{align}
where $C$ is a positive constant.
In other words, \emph{word2vec with the negative sampling is a gravity model}, with the mass given by the location's frequency $P(i)$, and the distance measured by their dot similarities.
While the gravity model generates the flow from the given mass and locations, {\it word2vec} finds the mass and locations that best explain the given trajectories.
Therefore, the embedding distance is inherently tied to the flow, and hence, has greater predictive power than the geographic distance.

@@

Negative sampling is a biased estimator~\autocite{Chia2010,Dyer2014} that optimizes a distinct model  from {\it word2vec}.
Negative sampling trains {\it word2vec} using a binary classification task as follows.
A center-context word pair $(i,j)$ is sampled from the given data as a positive pair, and $k$ negative pairs $(i,\ell)$ are sampled from a noise distribution $p_0(\ell)=P^\gamma(\ell)$. ($P(j)$ is the fraction of $j$ in the data, and $\gamma>0$ is a parameter).
Then, given a center-context word pair $(i,j)$, the model predicts whether it is a positive or negative sample using sigmoid function:
\begin{align}
	\label{eq:sigmoid}
	\frac{1}{1 + \exp\left( - \bm{u}_j \cdot \bm{v}_i \right)}.
\end{align}

This procedure does not guarantee that the embedding optimally converges, even when increasing the training samples and iterations~\autocite{Chia2010,Dyer2014}.
This problem is addressed by the noise contrastive estimation (NCE) \autocite{Chia2010}, which fits a probability model in the form:
\begin{align}
	\label{eq:nce}
	f(x) / \sum_{x'} f(x'),
\end{align}
where $f(x)$ is the unnormalized probability for an instance $x$.
As in negative sampling, NCE fits the model based on a binary classification task, with positive and negative samples randomly drawn from the given data and a noise distribution $p_0$, respectively, using a bias-corrected sigmoid function:
\begin{align}
	\label{eq:sigmoid2}
	\frac{1}{1 + \exp\left( -f(x) + \log p_0(x) \right)}.
\end{align}
In case of the {\it word2vec}, the instance is the dot similarity $x=\bm{u}_i \cdot \bm{v}_j$,  $f(x)=\exp(x)$, and $p_0(j)=P^\gamma(j)$.

@@

We model co-occurences $T_{ij}$ for locations $i$ and $j$ (referred to as flux), using the gravity law of mobility~\autocite{zipf1946gravity}.
The gravity law of mobility, which was inspired by Newton's law of gravity, postulates that attraction between two locations is a function of their population and the distance between them.
This formulation and variants have proven useful for modeling and predicting many kinds of mobility~\autocite{jung2008highwaygravity, curiel2018citygravity, truscott2012epidemicgravity, hong2016busgravity}.
In the gravity law of mobility, the \textit{expected flux}, $\hat{T}_{ij}$ between two locations $i$ and $j$ is defined as,

%
% Gravity law EQ
%
\begin{equation}
	\label{eq:gravity_basic}
	\hat{T}_{ij} = Cm_{i}m_{j}f(r_{ij}),
\end{equation}
where $m_{i}$ and $m_{j}$ are the population of locations, defined as the total number of passenger who passed through each airport for U.S. airport itineraries,  the total number of customer who booked with each accommodation for Korean accommodation reservations, and the yearly-average count of unique authors, both mobile and non-mobile, affiliated with each organization for scientific mobility.
$f(r_{ij})$ is a decay function of distance $r_{ij}$ between locations $i$ and $j$.
Here, we used the most basic gravity model which assumes symmetry of the flow $\hat{T}_{ij} =\hat{T}_{ji}$ and distance $r_{ij} = r_{ji}$, while there are four proposed variants \autocite{wilson2011entropy}.
There are two popular forms for the $f(r_{ij})$:
one is a power law function in the form $f(r_{ij})= r^{-\alpha}_{ij} \ (\alpha > 0)$, and the other is an exponential function in the form $f(r_{ij}) = e^{-\beta r_{ij}} \ (\beta > 0)$ \autocite{chen2015distance}.
The parameters for $f(r_{ij})$ and $C$ are fit to given mobility data using a log-linear regression~\autocite{jung2008highwaygravity, curiel2018citygravity, truscott2012epidemicgravity, hong2016busgravity, simini2012universal}.
